---
title: "ROY"
author: "Kevin O'Donnell"
date: '2022-08-28'
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(tidyverse)
library(nbastatR)
#library(nbaTools)
#library(NBAloveR)
#library(ncaahoopR)
#library(statsnbaR)
#library(hoopR)
#library(Hmisc)
library(corrplot)
library(reshape2)
library(ggrepel)
library(modelr)
library(knitr)
library(htmlTable)
library(stringr)
#library(pscl)
#library(boot)
library(caret)
#library(smotefamily)
#library(DMwR2)
library(ggthemes)
library(car)
```

# MVP 

## Data Collection, Merging, and Cleaning
```{r}
#MVP
mvp <- bref_awards(awards = c("Most Valuable Player"))
mvp <- mvp %>% select(namePlayer, slugSeason) %>% mutate(isMVP = 1)
mvp$namePlayer <- ifelse(mvp$namePlayer == "Nikola Jokić", "Nikola Jokic", mvp$namePlayer)
#player_stats <- teams_players_stats(seasons = 2021, types="player", tables=c("general", "defense"), season_types="Regular Season", modes="PerGame", is_pace_adjusted=T, measures="Base")

#award votes and award shares
voting <- read.csv("mvp_votings.csv")
voting <- voting %>% select(player, season, award_share)

#advanced and per game stats
player_stats <- bref_players_stats(seasons = 1954:2022, tables = c("advanced", "per_game"))
player_stats <- player_stats %>% group_by(namePlayer, slugSeason) %>% slice(1)

#combining mvp (Y/N) with player_stats
mvp_player_stats <- left_join(player_stats, voting, by=c("namePlayer"="player", "slugSeason"="season"))
mvp_player_stats <- mvp_player_stats %>% mutate(award_share = ifelse(is.na(award_share), 0, award_share))

mvp_player_stats2 <- left_join(mvp_player_stats, mvp, by=c("namePlayer", "slugSeason"))
mvp_player_stats2 <- mvp_player_stats2 %>% mutate(isMVP = ifelse(is.na(isMVP), 0, 1))

#what draft pick they were
#need more identifiers to prevent duplicates
drafts <- drafts(draft_years = 1952:2021, nest_data = FALSE, return_message = TRUE)
drafts <- drafts %>% select(namePlayer, numberPickOverall, idPlayer)

#duplicate players for some reason - find out how to filter for correct draft pick (not always first or last)
#might have to do manual
dups <- drafts %>% arrange(idPlayer) %>% group_by(idPlayer) %>% filter(n()>1)
dup_ids <- as.character(unique(dups$idPlayer))
dup_names <- as.character(unique(dups$namePlayer))

full_draft <- read.csv("NBA_Full_Draft_1947-2018.csv")
full_draft <- full_draft %>% select(Player, Year, Pick)
#filter for only names in dup_names vector
full_draft <- full_draft %>% filter(Player %in% dup_names)
draft_corrections <- left_join(dups, full_draft, by=c("namePlayer"="Player"))
draft_corrections <- draft_corrections %>% 
  group_by(namePlayer) %>% 
  filter(Year == max(Year)) %>%
  filter(!duplicated(namePlayer)) %>%
  select(namePlayer, Pick)

drafts2 <- left_join(drafts, draft_corrections, by="namePlayer")
drafts2 <- drafts2 %>% mutate(numberPickOverall = ifelse(is.na(Pick), numberPickOverall, Pick)) %>%
  unique() %>%
  select(-Pick)


player_stats_draft_pick <- left_join(mvp_player_stats2, drafts2, by=c("idPlayerNBA"="idPlayer", "namePlayer"))

#physical attributes
attributes <- read.csv("Players.csv")
attributes <- attributes[2:4]
attributes$Player <- str_replace(attributes$Player, '\\*', '')

#adding attirbutes to main data
full_player_stats <- left_join(player_stats_draft_pick, attributes, by=c("namePlayer"="Player"))

#checking for duplicates
nrow(full_player_stats %>% filter(yearSeason>1980) %>% group_by(namePlayer, yearSeason) %>% filter(n()>1))

#prior MVPs variables
full_player_stats <- full_player_stats %>% 
  group_by(namePlayer) %>% 
  arrange(yearSeason) %>% 
  mutate(totalMVP = cumsum(isMVP))

#removing unnecessary variables 
full_player_stats_clean <- full_player_stats %>% select(c("slugSeason", "namePlayer", "groupPosition", "yearSeason", 
"slugPosition", "agePlayer", "slugTeamBREF", "countGames", "slugTeamsBREF", "ratioPER", 
"pctTrueShooting", "pct3PRate", "pctFTRate", "pctORB", "pctDRB", 
"pctTRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", "ratioOWS", 
"ratioDWS", "ratioWS", "ratioWSPer48", "ratioOBPM", "ratioDBPM", 
"ratioBPM", "ratioVORP", "countGamesStarted", 
"pctFG", "pctFG3", "pctFG2", "pctEFG", "pctFT", "minutesPerGame", 
"fgmPerGame", "fgaPerGame", "fg3mPerGame", "fg3aPerGame", "fg2mPerGame", 
"fg2aPerGame", "ftmPerGame", "ftaPerGame", "orbPerGame", "drbPerGame", 
"trbPerGame", "astPerGame", "stlPerGame", "blkPerGame", "tovPerGame", 
"pfPerGame", "ptsPerGame", "numberPickOverall", "height", "weight", "isMVP", "totalMVP", "award_share"))

#lagging stats to use their previous year's stats, not their current year, so I can predict before season starts
lagged_stats <- full_player_stats_clean %>%
  group_by(namePlayer) %>%
  arrange(yearSeason) %>%
  mutate(award_share_lagged = award_share, across(c("countGames", "ratioPER", 
"pctTrueShooting", "pct3PRate", "pctFTRate", "pctORB", "pctDRB", 
"pctTRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", "ratioOWS", 
"ratioDWS", "ratioWS", "ratioWSPer48", "ratioOBPM", "ratioDBPM", 
"ratioBPM", "ratioVORP", "countGamesStarted", 
"pctFG", "pctFG3", "pctFG2", "pctEFG", "pctFT", "minutesPerGame", 
"fgmPerGame", "fgaPerGame", "fg3mPerGame", "fg3aPerGame", "fg2mPerGame", 
"fg2aPerGame", "ftmPerGame", "ftaPerGame", "orbPerGame", "drbPerGame", 
"trbPerGame", "astPerGame", "stlPerGame", "blkPerGame", "tovPerGame", 
"pfPerGame", "ptsPerGame", "isMVP", "totalMVP", "award_share_lagged"), ~ lag(x=.x, n=1)))

full_player_stats_clean %>% filter(namePlayer == "LeBron James")
lagged_stats %>% filter(namePlayer == "LeBron James")

team_stats <- data.frame()
for (i in 1980:2022) {
  df <- bref_teams_stats(seasons = i, assign_to_environment = F)
  df <- df[[2]][[1]]
  df <- df %>% select(slugSeason, yearSeason, slugTeamBREF, pctWins, ptsPerGameTeam, astPerGameTeam, drbPerGameTeam, tovPerGameTeam, pctFGPerGameTeam, pctFTPerGameTeam)
  team_stats <- rbind(team_stats, df)
}

#mess with year variable to allign player stats with team stats from prior year
team_stats <- team_stats %>% mutate(year = as.numeric(substr(slugSeason, 1, 4))+1, merge_year = year+1)

lagged_stats <- left_join(lagged_stats, team_stats, by=c("yearSeason"="merge_year", "slugTeamBREF"))

lagged_stats %>% filter(namePlayer == "LeBron James")

team_player_stats <- left_join(full_player_stats_clean, team_stats, by=c("slugTeamBREF", "yearSeason")) %>%
  filter(yearSeason > 1980)
```

# Initial Data Viz
```{r}
numeric <- dplyr::select(lagged_stats, where(is.numeric))
numeric <- numeric[,-1]
cormat <- cor(numeric, use = "complete.obs")
melted_cormat <- melt(cormat)
corrplot(cormat)
#correlation matrix (colorful)
ggplot(melted_cormat, aes(Var1, Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))

# Correlation matrix (1/3)
melted_cormat <- melted_cormat %>% arrange(Var1)

melted_cormat1 <- melted_cormat %>% filter(Var1 %in% c("yearSeason", "agePlayer", "countGames", 
"ratioPER", "pctTrueShooting", "pct3PRate", "pctFTRate", "pctORB", 
"pctDRB", "pctTRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", 
"ratioOWS", "ratioDWS", "ratioWS", "ratioWSPer48", "ratioOBPM"), Var2 %in% c("yearSeason", "agePlayer", "countGames", 
"ratioPER", "pctTrueShooting", "pct3PRate", "pctFTRate", "pctORB", 
"pctDRB", "pctTRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", 
"ratioOWS", "ratioDWS", "ratioWS", "ratioWSPer48", "ratioOBPM"))

melted_cormat2 <- melted_cormat %>% filter(Var1 %in% c("ratioDBPM", "ratioBPM", "ratioVORP", "countGamesStarted", "pctFG", "pctFG3", "pctFG2", "pctEFG", "pctFT", "minutesPerGame", "fgmPerGame", 
"fgaPerGame", "fg3mPerGame", "fg3aPerGame", "fg2mPerGame", "fg2aPerGame", 
"ftmPerGame", "ftaPerGame", "orbPerGame", "drbPerGame") , Var2 %in% c("ratioDBPM", "ratioBPM", "ratioVORP", "countGamesStarted", "pctFG", 
"pctFG3", "pctFG2", "pctEFG", "pctFT", "minutesPerGame", "fgmPerGame", "fgaPerGame", "fg3mPerGame", "fg3aPerGame","fg2mPerGame", "fg2aPerGame", "ftmPerGame", "ftaPerGame", "orbPerGame", "drbPerGame"))

melted_cormat3 <- melted_cormat %>% filter(Var1 %in% c("trbPerGame", "astPerGame", "stlPerGame", "blkPerGame", "tovPerGame", "pfPerGame", "ptsPerGame", "numberPickOverall", "height", "weight", "isMVP", "totalMVP", "yearSeason.y", "pctWins", "ptsPerGameTeam", "astPerGameTeam", "drbPerGameTeam", "tovPerGameTeam", "pctFGPerGameTeam", "pctFTPerGameTeam", 
"year") , Var2 %in% c("trbPerGame", "astPerGame", "stlPerGame", "blkPerGame", "tovPerGame", "pfPerGame", "ptsPerGame", "numberPickOverall", "height", "weight", "isMVP", "totalMVP", "yearSeason.y", "pctWins", "ptsPerGameTeam", "astPerGameTeam", "drbPerGameTeam", "tovPerGameTeam", "pctFGPerGameTeam", "pctFTPerGameTeam", 
"year"))

ggplot(melted_cormat1, aes(Var1, Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1)) +
  labs(title = "Correlations between Numeric Variables (Pt. 1)", x="Variable 1", y= "Variable 2") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"))


ggplot(melted_cormat2, aes(Var1, Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
  labs(title = "Correlations between Numeric Variables (Pt. 2)", x="Variable 1", y= "Variable 2") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"))


ggplot(melted_cormat3, aes(Var1, Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 12, hjust = 1))+
  labs(title = "Correlations between Numeric Variables (Pt. 3)", x="Variable 1", y= "Variable 2") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"))

#histograms of all numeric vars
hist.data.frame(numeric)

#percentage of values missing for each column
colSums(is.na(lagged_stats))/nrow(lagged_stats) 

#ugly plot but supposed to be every variable related to isMVP
gathered <- lagged_stats %>% gather(key= "variable", value = "value", -isMVP)
ggplot(gathered, aes(x = value, y = isMVP)) +
  geom_point() +
  facet_wrap(~variable)

lagged_stats %>% group_by(isMVP) %>%
  summarise(mean(agePlayer))

mvps <- full_player_stats_clean %>% filter(isMVP ==1)
bums <- full_player_stats_clean %>% filter(isMVP ==0)

mvps_1981 <- mvps %>% filter(yearSeason >= 1981)
bums_1981 <- bums %>% filter(yearSeason >= 1981)

#t-tests comparing mvps to non mvps for different stats
t.test(bums_1981$agePlayer, mvps_1981$agePlayer)
t.test(bums_1981$ratioPER, mvps_1981$ratioPER)
t.test(bums_1981$ratioWS, mvps_1981$ratioWS)
t.test(bums_1981$ptsPerGame, mvps_1981$ptsPerGame)
t.test(bums_1981$astPerGame, mvps_1981$astPerGame)
t.test(bums_1981$trbPerGame, mvps_1981$trbPerGame)
t.test(bums_1981$pctTrueShooting, mvps_1981$pctTrueShooting)
t.test(bums_1981$height, mvps_1981$height)
weight_t_test <- t.test(bums_1981$weight, mvps_1981$weight)

numeric_names <- names(dplyr::select(mvps_1981, where(is.numeric)))
numeric_names <- numeric_names[-c(1,2,52,53,54)]

p_values <- c()
greater_or_less <- c()
for (i in 1:49) {
  x <- bums_1981 %>% select(numeric_names[i])
  y <- mvps_1981 %>% select(numeric_names[i])
  p_values[i] <- t.test(x[2], y[2])$p.value
  greater_or_less[i] <- ifelse(t.test(x[2], y[2])$estimate[2] > t.test(x[2], y[2])$estimate[1], "Greater", "Less")
}
p_values <- data.frame(numeric_names, p_values, greater_or_less)
p_values <- p_values %>% mutate(significant = ifelse(p_values<0.05, "Yes", "No")) %>% arrange(p_values) %>% rename("Variable" = numeric_names, "P-Value" = p_values, "Significant at alpha=0.05?" = significant, "Is Mean for MVPs Greater or Less Than Mean for non-MVPs?" = greater_or_less)
p_values$`P-Value` <- round(p_values$`P-Value`, 5)
htmlTable(p_values)


#correlations of each variable with award_share
numeric <- dplyr::select(filter(full_player_stats_clean, yearSeason>1980), where(is.numeric))
numeric <- numeric %>% select(-yearSeason, -isMVP, -namePlayer)
numeric <- numeric[,-1]

cors <- c()
numeric_names <- c()
y <- numeric %>% select(award_share)
for (i in 1:50) {
  x <- numeric[,i]
  numeric_names[i] <- names(x)
  cors[i] <- cor(x,y, use="complete.obs")
}
cors_df <- data.frame(numeric_names, cors) 
cors_df <- cors_df %>% arrange(-cors) %>% rename("Variable" = numeric_names, "Correlation with Award Share" = cors)
cors_df$`Correlation with Award Share` <- round(cors_df$`Correlation with Award Share`, digit = 5)
htmlTable(cors_df)

cor(team_player_stats$award_share, team_player_stats$height, use="complete.obs")
cor(team_player_stats$award_share, team_player_stats$weight, use="complete.obs")


#relationships between variables and isMVP using simple logistic models
all <- full_player_stats_clean %>% select(-yearSeason, -namePlayer, -yearSeason, -slugTeamsBREF)
all <- all[,-1]

marginal_p_values <- c()
all_names <- c()
y <- all$isMVP
all <- all %>% select(-isMVP)
for (i in 1:55) {
  x <- all[,i]
  total <- cbind(y, x)
  all_names[i] <- names(x)
  marginal_p_values[i] <- summary(glm(y ~ ., data = total, family = "gaussian"))$coefficients[2,4]
}
marginal_p_values_df <- data.frame(all_names, marginal_p_values) %>% 
  arrange(marginal_p_values) %>%
  mutate(significant = ifelse(marginal_p_values<0.05, "Yes", "No")) %>%
  rename("Variable" = all_names, "P-Value" = marginal_p_values, "Significant Marginal Relationship at alpha=0.05?" = significant)
marginal_p_values_df <- marginal_p_values_df[-55,]
marginal_p_values_df$`P-Value` <- round(marginal_p_values_df$`P-Value`, 5)
htmlTable(marginal_p_values_df)

summary(glm(isMVP ~ pctWins, data = team_player_stats, family = "gaussian"))$coefficients[2,4]
summary(glm(isMVP ~ ptsPerGameTeam, data = team_player_stats, family = "gaussian"))$coefficients[2,4]
summary(glm(isMVP ~ astPerGameTeam, data = team_player_stats, family = "gaussian"))$coefficients[2,4]
summary(glm(isMVP ~ drbPerGameTeam, data = team_player_stats, family = "gaussian"))$coefficients[2,4]
summary(glm(isMVP ~ tovPerGameTeam, data = team_player_stats, family = "gaussian"))$coefficients[2,4]
summary(glm(isMVP ~ pctFGPerGameTeam, data = team_player_stats, family = "gaussian"))$coefficients[2,4]
summary(glm(isMVP ~ pctFTPerGameTeam, data = team_player_stats, family = "gaussian"))$coefficients[2,4]


#relationships between categorical variables and award_share / isMVP
categorical <- dplyr::select(filter(full_player_stats_clean, yearSeason>1980), where(is.character))
chisq.test(full_player_stats_clean$groupPosition, as.factor(full_player_stats_clean$isMVP))
full_player_stats_clean %>% group_by(groupPosition) %>% summarise(sum(isMVP))
chisq.test(full_player_stats_clean$slugPosition, as.factor(full_player_stats_clean$isMVP))
full_player_stats_clean %>% group_by(slugPosition) %>% summarise(sum(isMVP))

summary(aov(award_share ~ groupPosition, data=lagged_stats_model_likely))
summary(aov(award_share ~ slugPosition, data=lagged_stats_model_likely))


#MVPs by draft position
full_player_stats %>% filter(isMVP==1) %>% select(numberPickOverall)

#mvps by position
ggplot(mvps, aes(x=slugPosition)) +
  geom_bar() +
  labs(title = "MVPs by Position Since 1955")

mvps <- mvps %>% mutate(decade = paste0(as.character(yearSeason - yearSeason %% 10), "s"))
mvps_1981 <- mvps_1981 %>% mutate(decade = paste0(as.character(yearSeason - yearSeason %% 10), "s"))
ggplot(mvps_1981, aes(x=fct_relevel(slugPosition, "PG", "SG", "SF", "PF", "C"), fill=decade)) +
  geom_bar() +
  labs(title = "MVPs by Position Since 1980", x="Position", y="Number of MVPs", subtitle = "Segmented by Decade") +
  theme(plot.title = element_text(hjust = 0.5, color = "Purple"), plot.subtitle = element_text(hjust = 0.5, color = "Salmon")) +
  guides(fill=guide_legend(title="Decade"))


#DWS to OWS for MVPs (and maybe all just highlight MVPs)
lagged_stats_plot <- lagged_stats %>% mutate(factor = as.factor(ifelse(isMVP==1, "Yes", "No")))
full_player_stats_clean_plot <- full_player_stats_clean %>% 
  mutate(factor = as.factor(ifelse(isMVP==1, "Yes", "No"))) %>% 
  filter(yearSeason > 1980)

quantile(full_player_stats_clean_plot$ratioOWS, probs = 1-.1/100)
quantile(full_player_stats_clean_plot$ratioDWS, probs = 1-.1/100)

nrow(full_player_stats_clean_plot %>% filter(ratioOWS > 13.1988))
nrow(full_player_stats_clean_plot %>% filter(ratioDWS > 7.1988))

highlighted <- full_player_stats_clean_plot %>% filter(ratioOWS > 13.1988 | ratioDWS > 7.1988)
highlighted %>% filter(ratioOWS > 13.1988, isMVP==1)
highlighted %>% filter(ratioDWS > 7.1988, isMVP==1)

ggplot(full_player_stats_clean_plot, aes(x=ratioDWS, y=ratioOWS, color=factor)) +
  geom_point() +
  #geom_label_repel(data = filter(full_player_stats_clean_plot, (ratioOWS>14 | ratioDWS>8)), aes(label=namePlayer), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3) +
  labs(title = "Players Offensive Win Shares vs Their Defensive Win Shares", x="DWS", y="OWS", subtitle = "Since the 1980-81 Season") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"), plot.subtitle = element_text(hjust = 0.5, color = "#00BFC4")) +
  guides(color=guide_legend(title="MVP?")) +
  #geom_hline(yintercept=(quantile(full_player_stats_clean_plot$ratioOWS, probs = 1-.1/100)), color="grey") +
  #geom_vline(xintercept=(quantile(full_player_stats_clean_plot$ratioDWS, probs = 1-.1/100)), color="grey") +
  geom_point(data = highlighted, aes(x=ratioDWS, y=ratioOWS, color=factor), size=2.75, pch=21)
  


#FTM vs FTA
quantile(full_player_stats_clean_plot$ftmPerGame / full_player_stats_clean_plot$ftaPerGame, na.rm = T, probs = .75)
quantile(full_player_stats_clean_plot$ftaPerGame, probs =.75)
filter(full_player_stats_clean_plot, isMVP == 1, pctFT > 0.8181818)
filter(full_player_stats_clean_plot, isMVP == 1, ftaPerGame > 2.9)

ggplot(full_player_stats_clean_plot %>% arrange(factor), aes(x=ftaPerGame, y=ftmPerGame, color=factor)) +
  geom_point(aes(size=factor)) +
  scale_size_discrete(range = c(1, 4), guide='none') +
  #geom_label_repel(data = filter(full_player_stats_clean_plot, (ratioOWS>14 | ratioDWS>8)), aes(label=namePlayer), point.padding = unit(0.5, "lines"), box.padding = unit(0.5, "lines"), size=3) +
  labs(title = "Players Free Throws Made vs Free Throws Attempted Per Game", x="FTA", y="FTM", subtitle = "Since the 1980-81 Season") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"), plot.subtitle = element_text(hjust = 0.5, color = "#00BFC4"), panel.background = element_rect(fill = "white"), panel.grid.major = element_line(color = "lightgrey")) +
  guides(color=guide_legend(title="MVP?")) +
  geom_abline(slope = 0.8181818, color="black") +
  geom_hline(yintercept=quantile(full_player_stats_clean_plot$ftmPerGame, probs =.75, na.rm = T), color="black") +
  geom_vline(xintercept=quantile(full_player_stats_clean_plot$ftaPerGame, probs =.75, na.rm = T), color="black") 
  #geom_point(data = highlighted, aes(x=ratioDWS, y=ratioOWS, color=factor), size=2.75, pch=21)


#VORP vs Award Share
quantile(full_player_stats_clean_plot$ratioVORP, probs =.9985)


ggplot(filter(full_player_stats_clean_plot, yearSeason < 2018), aes(x=ratioVORP, y=award_share, color=factor)) +
  geom_point() +
  labs(title = "Award Share by VORP", x="VORP", y="Award Share", subtitle = "Since the 1980-81 Season") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"), plot.subtitle = element_text(hjust = 0.5, color = "#00BFC4"), legend.position = c(.9, .4)) +
  guides(color=guide_legend(title="MVP?")) +
  geom_label_repel(data = filter(full_player_stats_clean_plot, (ratioVORP < 4.5 & award_share > .5), yearSeason < 2018), aes(label=paste(namePlayer, slugSeason)), point.padding = unit(0.5, "lines"), box.padding = unit(.5, "lines"), size=3, nudge_x = -3, direction = "y") +
  coord_cartesian(clip = "off") +
  geom_label_repel(data = filter(full_player_stats_clean_plot, (ratioVORP > 11 | (ratioVORP > 9 & award_share < .25)), yearSeason < 2018), aes(label=paste(namePlayer, slugSeason)), point.padding = unit(0.5, "lines"), box.padding = unit(.5, "lines"), size=3, nudge_x = 2.5, direction = "y", xlim = c(-Inf, Inf), ylim = c(-Inf, Inf)) +
  expand_limits(x = 16)


#BLK% vs TRB%
ggplot(filter(full_player_stats_clean_plot, minutesPerGame>20), aes(x=pctTRB, y=pctBLK)) +
  geom_point() +
  labs(title = "Players Block Percentage (BLK%) vs Total Rebounding Percentage (TRB%)", y="BLK%", x="TRB%", subtitle = "Since the 1980-81 Season") +
  theme(plot.title = element_text(hjust = 0.5, color = "salmon"), plot.subtitle = element_text(hjust = 0.5, color = "#00BFC4")) +
  geom_smooth(method="lm")


#how many games played do MVPs have?
ggplot(mvps_1981, aes(x=countGamesStarted)) +
  geom_histogram()

ggplot(bums_1981, aes(x=countGamesStarted)) +
  geom_histogram()

#how many minutes do MVPs play?
ggplot(mvps_1981, aes(x=minutesPerGame)) +
  geom_histogram()

ggplot(bums_1981, aes(x=minutesPerGame)) +
  geom_histogram()

ggplot(mvps, aes(agePlayer)) +
  geom_histogram(binwidth = 1, color="Salmon", fill="black") +
  labs(title = "Age of MVPs", subtitle = "1954-Present", x="Age", y="Count") +
  theme(plot.title = element_text(hjust = 0.5, color = "Purple"), plot.subtitle = element_text(hjust = 0.5, color = "Salmon"))

#voting history
voting %>% distinct(season)
nrow(voting)/nrow(voting %>% distinct(season))
```

# Prepping model and modeling next year stats for feature creation 
```{r}
lagged_stats$slugPosition <- ifelse(lagged_stats$slugPosition == "C-F", "F-C", lagged_stats$slugPosition)

na_count <-sapply(lagged_stats, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

filter(lagged_stats, is.na(ptsPerGameTeam) & yearSeason>1981)

lagged_stats_model <- lagged_stats %>% filter(yearSeason>=1981) %>%
  mutate_at(vars(pctWins:pctFTPerGameTeam),~ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x))

na_count <-sapply(lagged_stats_model, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

write.csv(lagged_stats_model, "test.csv")
```

# More data filtering to only include players likely to get MVP
```{r}
mean(mvps_1981$minutesPerGame) - 3*sd(mvps_1981$minutesPerGame)
mean(mvps_1981$countGamesStarted) - 3*sd(mvps_1981$countGamesStarted)
mean(mvps_1981$countGames) - 3*sd(mvps_1981$countGames)

lagged_stats_model %>% filter(lead(isMVP, 1)==1) %>% filter(lead(minutesPerGame, 1) < mean(mvps_1981$minutesPerGame) - 3*sd(mvps_1981$minutesPerGame) | lead(countGamesStarted, 1) < mean(mvps_1981$countGamesStarted) - 3*sd(mvps_1981$countGamesStarted) | lead(countGames, 1)< mean(mvps_1981$countGames) - 3*sd(mvps_1981$countGames))

#filtering out wrong people - adjust for lag
#also dependent on mvps dataset which was run in the other markdown file but I haven't changed the code in this one to allow it
lagged_stats_model_likely2 <- lagged_stats_model %>% 
  group_by(namePlayer) %>%
  arrange(yearSeason) %>%
  filter(lead(minutesPerGame, 1) > (mean(mvps_1981$minutesPerGame) - 3*sd(mvps_1981$minutesPerGame))) %>%
  filter(lead(countGamesStarted, 1) > (mean(mvps_1981$countGamesStarted) - 3*sd(mvps_1981$countGamesStarted)))#%>%
  #filter(lead(countGames, 1) > (mean(mvps_1981$countGames) - 3*sd(mvps_1981$countGames)))

na_count <-sapply(lagged_stats_model_likely, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

lagged_stats_model_likely <- lagged_stats_model_likely %>% filter(!is.na(award_share_lagged))

na_count <-sapply(lagged_stats_model_likely, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

inner_join(lagged_stats_model_likely, mvps_1981, by=c("namePlayer", "yearSeason"))
lagged_stats_model %>% filter(namePlayer %in% c("Magic Johnson", "Michael Jordan")) %>% summarise(slugSeason.x, countGames, countGamesStarted, minutesPerGame)
```


# Initial Models
```{r}
lagged_stats_model_likely_aws <- lagged_stats_model_likely %>% filter(yearSeason<=2018)

train <- lagged_stats_model_likely_aws %>% filter(yearSeason<=2007)
test <- lagged_stats_model_likely_aws %>% filter(yearSeason>2007)

#using all data points instead of just likely ones
lagged_stats_model_aws <- lagged_stats_model %>% filter(yearSeason<=2018)

train <- lagged_stats_model_aws %>% filter(yearSeason<=2008)
test <- lagged_stats_model_aws %>% filter(yearSeason>2008)


full_lm <- lm(award_share ~  factor(groupPosition) + 
 factor(slugPosition) +  agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam + award_share_lagged + isMVP + totalMVP, data = train)

summary(full_lm)

empty_lm <- lm(award_share ~  1, data = na.omit(train))

#backward selection
stepAIC(full_lm, direction = "backward")
stepAIC(empty_lm, direction = "forward", scope=list(lower=empty_lm, upper=~factor(groupPosition) + 
 factor(slugPosition) +  agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam + award_share_lagged + isMVP + totalMVP))

forward_select <- lm(formula = award_share ~ award_share_lagged + ratioVORP + fg3mPerGame + 
    ratioWS + ftaPerGame + stlPerGame + drbPerGame + minutesPerGame + 
    pctFTRate + factor(groupPosition) + fg2aPerGame + pctUSG + 
    ratioBPM + ptsPerGameTeam + astPerGame + pctDRB + pctAST + 
    agePlayer + pctFGPerGameTeam + tovPerGameTeam + ftmPerGame + 
    pctFT + ratioOWS, data = na.omit(train))
vif(forward_select)

backward_select <- lm(formula = award_share ~ factor(groupPosition) + ratioPER + pctTrueShooting+ pctFTRate + pctDRB + pctAST + pctSTL + pctBLK + pctTOV + pctUSG + ratioOWS + ratioOBPM + ratioDBPM + ratioVORP + countGamesStarted + pctFG + pctEFG + pctFT + minutesPerGame + fgmPerGame + ftmPerGame + drbPerGame + stlPerGame + trbPerGame + pfPerGame + ptsPerGame + pctFGPerGameTeam + pctFTPerGameTeam + ptsPerGameTeam, data = train)

#choosing this one for now
backward_select <- lm(formula = award_share ~ agePlayer + pct3PRate + pctFTRate + 
    pctDRB + pctAST + pctUSG + ratioOWS + ratioWS + ratioDBPM + 
    ratioBPM + ratioVORP + pctFT + minutesPerGame + fgmPerGame + 
    fg3mPerGame + fg2aPerGame + ftaPerGame + drbPerGame + astPerGame + 
    ptsPerGame + ptsPerGameTeam + tovPerGameTeam + pctFGPerGameTeam, data = train)

#reducing multicolinearity
backward_select <- lm(formula = award_share ~ factor(groupPosition) + pctTrueShooting + pctFTRate + pctDRB + pctAST + pctSTL + pctBLK + pctTOV + pctUSG + ratioVORP + countGamesStarted + ftmPerGame + pfPerGame + pctFGPerGameTeam + ptsPerGameTeam, data = train)

backward_select <- lm(formula = award_share ~ agePlayer + pct3PRate + pctFTRate + 
    pctDRB + pctAST + pctUSG + 
    ratioBPM + ratioVORP + pctFT + ftaPerGame + drbPerGame + 
    ptsPerGameTeam + tovPerGameTeam + pctFGPerGameTeam + 
    award_share_lagged, data = train)

train_numeric <- numeric <- dplyr::select(train, where(is.numeric))
train_numeric <- train_numeric[,-1]
cor <- cor(train_numeric, use = "complete.obs")
cor <- as.data.frame(cor)
cor <- cor %>% select(c("ratioPER", "pctTrueShooting", "pctFTRate", "pctDRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", "ratioOWS", "ratioOBPM", "ratioDBPM", "ratioVORP", "countGamesStarted", "pctFG", "pctEFG", "pctFT", "minutesPerGame", "fgmPerGame", "ftmPerGame", "drbPerGame", "stlPerGame", "trbPerGame", "pfPerGame", "ptsPerGame", "pctFGPerGameTeam", "pctFTPerGameTeam", "ptsPerGameTeam"))
cor <- cor[rownames(cor) %in% c("ratioPER", "pctTrueShooting", "pctFTRate", "pctDRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", "ratioOWS", "ratioOBPM", "ratioDBPM", "ratioVORP", "countGamesStarted", "pctFG", "pctEFG", "pctFT", "minutesPerGame", "fgmPerGame", "ftmPerGame", "drbPerGame", "stlPerGame", "trbPerGame", "pfPerGame", "ptsPerGame", "pctFGPerGameTeam", "pctFTPerGameTeam", "ptsPerGameTeam"),]

summary(backward_select)

#sqrt transformation
sqrt_lagged_model_likely <- train %>% mutate(sqrt_award_share = sqrt(award_share))

sqrt_lm <- lm(sqrt_award_share ~  factor(groupPosition) + 
 factor(slugPosition) +  agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam + award_share_lagged + isMVP + totalMVP, data = sqrt_lagged_model_likely)

summary(sqrt_lm)

stepAIC(sqrt_lm, direction = "backward")
sqrt_backward_select <- lm(formula = sqrt_award_share ~ agePlayer + ratioPER + pctTrueShooting + 
    pctFTRate + pctDRB + pctTRB + pctAST + pctUSG + ratioWSPer48 + 
    ratioOBPM + ratioVORP + countGamesStarted + pctFG + pctEFG + 
    pctFT + minutesPerGame + fg3mPerGame + fg2mPerGame + ftaPerGame + 
    drbPerGame + astPerGame + stlPerGame + blkPerGame + pfPerGame + 
    ptsPerGame + ptsPerGameTeam + tovPerGameTeam + pctFGPerGameTeam + 
    award_share_lagged + isMVP + totalMVP, data = sqrt_lagged_model_likely)


lagged_stats_pred <- add_predictions(test, backward_select, var="pred")
lagged_stats_pred %>% filter(is.na(pred)) %>% group_by(pred) %>% summarise(n())

#checking if model would have gotten it right in each year
lagged_stats_pred %>% filter(namePlayer == "Karl Malone")
lagged_stats_model_likely %>% filter(namePlayer == "Karl Malone")
lagged_stats_pred %>% filter(namePlayer == "Karl Malone") %>% summarise(yearSeason, slugSeason.x, minutesPerGame, countGames, countGamesStarted)
mean(mvps_1981$minutesPerGame) - 3*sd(mvps_1981$minutesPerGame)
mean(mvps_1981$countGamesStarted) - 3*sd(mvps_1981$countGamesStarted)
mean(mvps_1981$countGames) - 3*sd(mvps_1981$countGames)

#identifying duplicates
nrow(lagged_stats_model_likely %>% group_by(namePlayer, yearSeason) %>% filter(n()>1))


top_actual <- lagged_stats_model %>% 
  group_by(yearSeason) %>%
  top_n(1, award_share) %>% 
  summarise(namePlayer, slugSeason.x, award_share) %>%
  filter(yearSeason <= 2018)

top_predicted <- lagged_stats_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred, pred_squared = pred^2) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

```

# Checking Assumptions for Initial Model(s)
```{r}
#sqrt model

#no autocollinearity
library(car)
durbinWatsonTest(sqrt_lm)

#serious fanning - no constant variance
plot(sqrt_lm$fitted.values, sqrt_lm$residuals)
abline(h=0)

#residuals definitely not normal
qqnorm(sqrt_lm$residuals)
qqline(sqrt_lm$residuals)

#sqrt model (backward selection)

#no autocollinearity
durbinWatsonTest(sqrt_backward_select)

#serious fanning - no constant variance
plot(sqrt_backward_select$fitted.values, sqrt_backward_select$residuals)
abline(h=0)

#residuals definitely not normal
qqnorm(sqrt_backward_select$residuals)
qqline(sqrt_backward_select$residuals)


#normal linear model (backward selection)

#no autocollinearity
durbinWatsonTest(backward_select)

#serious fanning - no constant variance
plot(backward_select$fitted.values, backward_select$residuals)
abline(h=0)

#residuals definitely not normal
qqnorm(backward_select$residuals)
qqline(backward_select$residuals)


#normal linear model (full)

#no autocollinearity
durbinWatsonTest(full_lm)

#serious fanning - no constant variance
plot(full_lm$fitted.values, full_lm$residuals)
abline(h=0)

#residuals definitely not normal
qqnorm(full_lm$residuals)
qqline(full_lm$residuals)


#summarize all
plot(sqrt_lm)
plot(sqrt_backward_select)
plot(full_lm)
plot(backward_select)

vif(sqrt_lm)
vif(sqrt_backward_select)
vif(full_lm)
vif(backward_select)
```

# Basic vs Advanced Linear Models
```{r}
#just basic stats
basic_lm <- lm(award_share ~ 
 pctTrueShooting +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  + fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame, data = lagged_stats_model_likely)

#just advanced stats 
adv_lm <- lm(award_share ~
 ratioPER   +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP, data = lagged_stats_model_likely)

summary(basic_lm)
summary(adv_lm)
```

# Models for Award Share - Only MVPs
```{r}
full_lm_mvps_only <- lm(award_share ~  factor(groupPosition) + 
 factor(slugPosition) +  agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam + award_share_lagged + isMVP + totalMVP, data = filter(lagged_stats_model_likely, award_share>0))

summary(full_lm_mvps_only)

stepAIC(full_lm_mvps_only, direction = "backward")
backward_select_mvps_only <- lm(formula = award_share ~ agePlayer + countGames + pctTrueShooting + 
    pctFTRate + pctTOV + pctUSG + ratioWS + pctFG2 + pctFT + 
    fg3aPerGame + fg2mPerGame + drbPerGame + pfPerGame + 
    drbPerGameTeam + pctFGPerGameTeam + award_share_lagged, data = filter(lagged_stats_model_likely, 
    award_share > 0))

plot(backward_select_mvps_only)
plot(full_lm_mvps_only)
```

# Down/Up sampling 
```{r}
#percent of positive award share and 0 award share in the data (all from 1980 onward)
lagged_stats_model %>% mutate(class1 = ifelse(award_share>0, 1, 0)) %>% group_by(class1) %>% summarise(n()/nrow(lagged_stats_model))

#sample <- sample(c(TRUE, FALSE), nrow(lagged_stats_model), replace=TRUE, prob=c(0.7,0.3))
#train  <- lagged_stats_model[sample, ]
#test   <- lagged_stats_model[!sample, ] 

lagged_stats_model_aws <- lagged_stats_model %>% filter(yearSeason<=2018)

train <- lagged_stats_model_aws %>% filter(yearSeason<=2008)
test <- lagged_stats_model_aws %>% filter(yearSeason>2008)

predictors <- train %>% select(-award_share)
award_share_df <- train %>% select(yearSeason, award_share)

class_membership <- train %>% mutate(class1 = ifelse(award_share>0, 1, 0)) 
class_membership <- class_membership[,71]
class_membership$class1 <- as.factor(class_membership$class1)
downSampled <- downSample(predictors, class_membership$class1, list = FALSE, yname = "Class")
 
downSampled %>% group_by(Class) %>% summarise(n()/nrow(downSampled))
downSampled <- left_join(downSampled, award_share_df, by=c("namePlayer", "yearSeason"))


down_sampled_lm <- lm(award_share ~  factor(groupPosition) + 
 agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam + totalMVP, data = downSampled)

stepAIC(down_sampled_lm, direction = "backward")

down_sampled_lm <- lm(formula = award_share ~ agePlayer + 
    ratioPER + pctTrueShooting + ratioOWS + ratioWS + ratioWSPer48 + 
    ratioVORP + pctEFG + pctFT + minutesPerGame + fgaPerGame + 
    fg3mPerGame + fg3aPerGame + ftaPerGame + drbPerGame + astPerGame + 
    blkPerGame + pfPerGame + pctWins + ptsPerGameTeam + drbPerGameTeam + 
    totalMVP, data = downSampled)
summary(down_sampled_lm)

down_sampled_pred <- add_predictions(test, down_sampled_lm, var="pred")

#find different way to evaluate accuracy - find top actual award share within test data (not going to be actual MVP but it makes more sense I think)
#also go back and split test and train by year instead of randomly - can actually evaluate true MVP winners instead
top_actual <- lagged_stats_model %>% 
  group_by(yearSeason) %>%
  top_n(1, award_share) %>% 
  summarise(namePlayer, slugSeason.x, award_share) %>%
  filter(yearSeason <= 2018)

top_predicted <- down_sampled_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

```

# Lagged GLMs (full, backward selection, basic, advanced)
```{r}
glm_train <- train %>%  
  group_by(namePlayer) %>%
  arrange(yearSeason) %>%
  mutate(isMVP = lead(isMVP, 1)) %>%
  mutate(isMVP = ifelse(is.na(isMVP), 0, isMVP))

glm_test <- test %>% 
  group_by(namePlayer) %>%
  arrange(yearSeason) %>%
  mutate(isMVP = lead(isMVP, 1)) %>%
  mutate(isMVP = ifelse(is.na(isMVP), 0, isMVP)) %>%
  mutate(isMVP = ifelse(namePlayer=="James Harden" & yearSeason==2018, 1, isMVP))

glm_test %>% filter(is.na(isMVP))

full_glm <- glm(isMVP ~  factor(groupPosition) + 
 factor(slugPosition) +  agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam, data = glm_train, family = binomial)

summary(full_glm)

empty_glm <- glm(isMVP ~ 1, data = glm_train, family=binomial)

stepAIC(full_glm, direction = "backward")

backward_select_glm <- glm(formula = isMVP ~ agePlayer + pct3PRate + pctFTRate + 
    pctDRB + pctAST + pctUSG + ratioOWS + ratioWS + ratioDBPM + 
    ratioBPM + ratioVORP + pctFT + minutesPerGame + fgmPerGame + 
    fg3mPerGame + fg2aPerGame + ftaPerGame + drbPerGame + astPerGame + 
    ptsPerGame + ptsPerGameTeam + tovPerGameTeam + pctFGPerGameTeam, 
    family = binomial, data = glm_train)

summary(backward_select_glm)

basic_glm <- glm(isMVP ~ 
 pctTrueShooting +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  + fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame, data = glm_train, family = binomial)

#just advanced stats 
adv_glm <- glm(isMVP ~
 ratioPER   +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP, data = lagged_stats_model_likely, family = binomial)

summary(basic_glm)
summary(adv_glm)

#confusion matrix and AIC
basic_glm$aic
adv_glm$aic

pred = predict(adv_glm, newdata=glm_test)
pred = ifelse(pred > 0.5, 1, 0)

confusionMatrix(data=as.factor(pred), as.factor(glm_test$isMVP), positive = "1")

#checking accuracy
lagged_stats_pred <- add_predictions(glm_test, adv_glm, var="pred", type = "response")
lagged_stats_pred %>% filter(is.na(pred)) %>% group_by(pred) %>% summarise(n())

top_predicted <- lagged_stats_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred, pred_squared = pred^2) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

```



# Trying current year model (not using prior year's stats)
redo without filtering because filtering excludes some MVPs
```{r}
current_stats <- team_player_stats %>% mutate(totalMVP = lag(totalMVP, n=1)) %>% filter(yearSeason <= 2018)
current_stats <- current_stats %>% 
  mutate_at(vars(pctWins:pctFTPerGameTeam),~ifelse(is.na(.x), mean(.x, na.rm = TRUE), .x)) %>%
  mutate(totalMVP = ifelse(is.na(totalMVP), 0, totalMVP))

current_stats_likely <- current_stats %>% 
  filter(minutesPerGame > (mean(mvps_1981$minutesPerGame) - 3*sd(mvps_1981$minutesPerGame))) %>%
  filter(countGamesStarted > (mean(mvps_1981$countGamesStarted) - 3*sd(mvps_1981$countGamesStarted)))%>%
  filter(countGames > (mean(mvps_1981$countGames) - 3*sd(mvps_1981$countGames)))

na_count <-sapply(current_stats_likely, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

#note: split is 2008 for unfiltered data
current_train <- current_stats_likely %>% filter(yearSeason<=2007)
current_test <- current_stats_likely %>% filter(yearSeason>2007)

current_train <- current_stats %>% filter(yearSeason<=2008)
current_test <- current_stats %>% filter(yearSeason>2008)

# checking initial class split
current_train %>% mutate(class1 = ifelse(award_share>0, 1, 0)) %>% group_by(class1) %>% summarise(n()/nrow(current_train))

# LM (no downsampling)
current_lm <- lm(award_share ~  factor(groupPosition) + 
 agePlayer +  countGames + 
 ratioPER  +  pctTrueShooting  +  pct3PRate  +  pctFTRate  +  pctORB  + 
 pctDRB  +  pctTRB  +  pctAST  +  pctSTL  +  pctBLK  +  pctTOV  +  pctUSG  + 
 ratioOWS  +  ratioDWS  +  ratioWS  +  ratioWSPer48  +  ratioOBPM  + 
 ratioDBPM  +  ratioBPM  +  ratioVORP  +  countGamesStarted +  pctFG  + 
 pctFG3  +  pctFG2  +  pctEFG  +  pctFT  +  minutesPerGame +  fgmPerGame  + 
 fgaPerGame  +  fg3mPerGame  +  fg3aPerGame  +  fg2mPerGame  +  fg2aPerGame  + 
 ftmPerGame  +  ftaPerGame  +  orbPerGame  +  drbPerGame  +  trbPerGame  + 
 astPerGame  +  stlPerGame  +  blkPerGame  +  tovPerGame  +  pfPerGame  + 
 ptsPerGame + pctWins + ptsPerGameTeam + astPerGameTeam + ptsPerGameTeam + drbPerGameTeam + tovPerGameTeam + pctFGPerGameTeam +  pctFTPerGameTeam + totalMVP, data = current_train)

stepAIC(current_lm, direction = "backward")

#totalMVP was taking over - removed it 
current_lm <- lm(formula = award_share ~ factor(groupPosition) + ratioPER + 
    pctTrueShooting + pctFTRate + pctDRB + pctAST + pctSTL + 
    pctBLK + pctTOV + pctUSG + ratioOWS + ratioOBPM + ratioDBPM + 
    ratioVORP + countGamesStarted + pctFG + pctEFG + pctFT + 
    minutesPerGame + fgmPerGame + fgaPerGame + fg3aPerGame + 
    fg2mPerGame + fg2aPerGame + ftmPerGame + ftaPerGame + drbPerGame + 
    trbPerGame + stlPerGame + blkPerGame + tovPerGame + pfPerGame + 
    ptsPerGameTeam + pctFGPerGameTeam + pctFTPerGameTeam, 
    data = current_train)
summary(current_lm)

#this is the one for now
current_lm <- lm(formula = award_share ~ agePlayer + pct3PRate + pctFTRate + 
    pctDRB + pctAST + pctUSG + ratioOWS + ratioWS + ratioDBPM + 
    ratioBPM + ratioVORP + pctFT + minutesPerGame + fgmPerGame + 
    fg3mPerGame + fg2aPerGame + ftaPerGame + drbPerGame + astPerGame + 
    ptsPerGame + ptsPerGameTeam + tovPerGameTeam + pctFGPerGameTeam, data = current_train)
summary(current_lm)
plot(current_lm)

current_pred <- add_predictions(current_test, current_lm, var="pred")

top_predicted <- current_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

sqrt(mean(current_lm$residuals^2))
sqrt(mean(full_lm$residuals^2))

# GLM (no downsampling)

current_glm <- glm(formula = isMVP ~ factor(groupPosition) + ratioPER + 
    pctTrueShooting + pctFTRate + pctDRB + pctAST + pctSTL + 
    pctBLK + pctTOV + pctUSG + ratioOWS + ratioOBPM + ratioDBPM + 
    ratioVORP + countGamesStarted + pctFG + pctEFG + pctFT + 
    minutesPerGame + fgmPerGame + fgaPerGame + fg3aPerGame + 
    fg2mPerGame + fg2aPerGame + ftmPerGame + ftaPerGame + drbPerGame + 
    trbPerGame + stlPerGame + blkPerGame + tovPerGame + pfPerGame + 
    ptsPerGameTeam + pctFGPerGameTeam + pctFTPerGameTeam, 
    data = current_train, family = binomial)

current_glm <- glm(formula = isMVP ~ pctTrueShooting + pctFTRate + pctBLK + 
    ratioOWS + ratioOBPM + ratioDBPM + ratioVORP + countGamesStarted + 
    pctEFG + fgmPerGame + fgaPerGame + fg3aPerGame + fg2mPerGame + 
    fg2aPerGame + ftmPerGame + ftaPerGame + trbPerGame + stlPerGame + 
    tovPerGame + pfPerGame + pctFGPerGameTeam, family = binomial, 
    data = current_train)
summary(current_glm)

#this is the one for now
current_glm <- glm(formula = isMVP ~ agePlayer + pct3PRate + pctFTRate + 
    pctDRB + pctAST + pctUSG + ratioOWS + ratioWS + ratioDBPM + 
    ratioBPM + ratioVORP + pctFT + minutesPerGame + fgmPerGame + 
    fg3mPerGame + fg2aPerGame + ftaPerGame + drbPerGame + astPerGame + 
    ptsPerGame + ptsPerGameTeam + tovPerGameTeam + pctFGPerGameTeam, family = binomial, 
    data = current_train)
summary(current_glm)

pred = predict(current_glm, newdata=current_test)
pred = ifelse(pred > 0.5, 1, 0)

confusionMatrix(data=as.factor(pred), as.factor(current_test$isMVP), positive = "1")

current_pred <- add_predictions(current_test, current_glm, var="pred", type = "response")

top_predicted <- current_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

# LM (with down-sampling/up-sampling)
predictors <- current_train %>% select(-award_share)
award_share_df <- current_train %>% select(yearSeason, award_share)

class_membership <- current_train %>% mutate(class1 = ifelse(award_share>0, 1, 0)) 
class_membership <- class_membership[,70]
class_membership$class1 <- as.factor(class_membership$class1)
upSampled <- upSample(predictors, class_membership$class1, list = FALSE, yname = "Class")

upSampled %>% group_by(Class) %>% summarise(n()/nrow(upSampled))
upSampled <- left_join(upSampled, award_share_df, by=c("namePlayer", "yearSeason"))

current_lm_ds <- lm(formula = award_share ~ factor(groupPosition) + ratioPER + 
    pctTrueShooting + pctFTRate + pctDRB + pctAST + pctSTL + 
    pctBLK + pctTOV + pctUSG + ratioOWS + ratioOBPM + ratioDBPM + 
    ratioVORP + countGamesStarted + pctFG + pctEFG + pctFT + 
    minutesPerGame + fgmPerGame + fgaPerGame + fg3aPerGame + 
    fg2mPerGame + fg2aPerGame + ftmPerGame + ftaPerGame + drbPerGame + 
    trbPerGame + stlPerGame + blkPerGame + tovPerGame + pfPerGame + 
    ptsPerGameTeam + pctFGPerGameTeam + pctFTPerGameTeam, 
    data = upSampled)
summary(current_lm_ds)
plot(current_lm_ds)

current_pred_ds <- add_predictions(current_test, current_lm_ds, var="pred")

top_predicted <- current_pred_ds %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)


# GLM (will have to upsample for this because there are too few observations
predictors <- current_train %>% select(-isMVP)
award_share_df <- current_train %>% select(yearSeason, isMVP)

class_membership <- current_train %>% mutate(class1 = ifelse(isMVP==1, 1, 0)) 
class_membership <- class_membership[,70]
class_membership$class1 <- as.factor(class_membership$class1)
upSampled <- upSample(predictors, class_membership$class1, list = FALSE, yname = "Class")
 
upSampled %>% group_by(Class) %>% summarise(n()/nrow(upSampled))
upSampled <- left_join(upSampled, award_share_df, by=c("namePlayer", "yearSeason"))

current_glm <-  glm(formula = isMVP ~ pctTrueShooting + pctFTRate + pctBLK + 
    ratioOWS + ratioOBPM + ratioDBPM + ratioVORP + countGamesStarted + 
    pctEFG + fgmPerGame + fgaPerGame + fg3aPerGame + fg2mPerGame + 
    fg2aPerGame + ftmPerGame + ftaPerGame + trbPerGame + stlPerGame + 
    tovPerGame + pfPerGame + pctFGPerGameTeam, family = binomial, 
    data = upSampled)
summary(current_glm)

current_pred <- add_predictions(current_test, current_glm, var="pred", type = "response")
current_pred %>% summarise(namePlayer, yearSeason, pred, award_share, isMVP)

top_predicted <- current_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

```

# SMOTE with current data
```{r}
current_train_numeric <- current_train %>% select_if(is.numeric)
current_train_numeric <- current_train_numeric[,-c(1,48)]
na_count <-sapply(current_train_numeric, function(y) sum(length(which(is.na(y)))))
current_train_numeric <- current_train_numeric %>% filter(complete.cases(current_train_numeric))

smote <- SMOTE(award_share ~ ., current_train_numeric, K=5)

smote <- SMOTE(X = current_train_numeric[,-53], target = current_train_numeric$award_share, K=5)
```

# Simulate resampling to see changes in the final model
problem: too few observations to converge?
```{r}
# past glm
sim_train1 <- glm_train %>% filter(isMVP==1)
sim_train0 <- glm_train %>% filter(isMVP==0) 
sim_train0 <- sim_train0[sample(1:nrow(sim_train0), 27, replace=F), ]
sim_train <- rbind(sim_train0, sim_train1)

simmed_glm <- glm(formula = isMVP ~ totalMVP + award_share_lagged + ratioDWS + ratioOWS + ratioVORP, 
    family = binomial, data = sim_train)

summary(simmed_glm)

#current glm
sim_train1 <- current_train %>% filter(isMVP==1)
sim_train0 <- current_train %>% filter(isMVP==0) 
sim_train0 <- sim_train0[sample(1:nrow(sim_train0), 26, replace=F), ]
sim_train <- rbind(sim_train0, sim_train1)

simmed_glm <- glm(formula = isMVP ~ pctTrueShooting + pctFTRate + pctBLK + 
    ratioOWS + ratioOBPM + ratioDBPM + ratioVORP + countGamesStarted + 
    pctEFG + fgmPerGame + fgaPerGame + fg3aPerGame + fg2mPerGame + 
    fg2aPerGame + ftmPerGame + ftaPerGame + trbPerGame + stlPerGame + 
    tovPerGame + pfPerGame + pctFGPerGameTeam, family = binomial, 
    data = sim_train)

summary(simmed_glm)

#past glm upsampling manually 
#gets similar predictions every time random sample is taken (good sign)
sim_train1 <- glm_train %>% filter(isMVP==1)
sim_train0 <- glm_train %>% filter(isMVP==0)
sim_train1 <- sim_train1[sample(1:nrow(sim_train1), 11022, replace=T), ]
sim_train1 %>% group_by(namePlayer) %>% summarise(n())
sim_train <- rbind(sim_train0, sim_train1)
sim_train <- sim_train %>% mutate(totalMVP = ifelse(is.na(totalMVP), 0, totalMVP))
simmed_glm <- glm(formula = isMVP ~ totalMVP + award_share_lagged + ratioDWS + ratioOWS + ratioVORP, 
    family = binomial, data = sim_train)

summary(simmed_glm)

lagged_stats_pred <- add_predictions(glm_test, simmed_glm, var="pred", type = "response")
lagged_stats_pred %>% summarise(namePlayer, yearSeason, pred, award_share, isMVP)

top_predicted <- lagged_stats_pred %>% 
  group_by(yearSeason) %>%
  top_n(1, pred) %>% 
  summarise(namePlayer, slugSeason.x, pred) %>% 
  filter(yearSeason <= 2018)

evaluation <- inner_join(top_actual, top_predicted, by=c("yearSeason", "namePlayer", "slugSeason.x"))
nrow(evaluation)/nrow(top_predicted)

pred = predict(simmed_glm, newdata=glm_test, type = "response")
pred = ifelse(pred > 0.5, 1, 0)

confusionMatrix(data=as.factor(pred), as.factor(glm_test$isMVP), positive = "1")
mean(pred == glm_test$isMVP, na.rm = T)
```

# XGBoost (classification)

## Training, validation, and test set creation
```{r}
set.seed(1)
xgboost_data <- current_stats
label = xgboost_data$isMVP
xgboost_data$isMVP = NULL
n = nrow(xgboost_data)
year = (xgboost_data %>% filter(yearSeason>2008))$yearSeason
year2 = (xgboost_data %>% filter(yearSeason>2002 & yearSeason<=2008))$yearSeason

names_backward_select <- c("agePlayer", "pct3PRate", "pctFTRate", "pctDRB", "pctAST", "pctUSG", 
    "ratioOWS", "ratioWS", "ratioBPM", "ratioDBPM","ratioVORP", "pctFT", "minutesPerGame", 
    "fgmPerGame", "fg3mPerGame", "fg2aPerGame", "ftaPerGame", "drbPerGame", "astPerGame", 
    "ptsPerGame", "ptsPerGameTeam", "tovPerGameTeam", "pctFGPerGameTeam")


train.data = data.matrix(xgboost_data %>% filter(yearSeason<=2002) %>% select(numeric_names) %>% select_if(is.numeric))
train.data <- train.data[,-1]
train.label = (current_train %>% filter(yearSeason<=2002))$isMVP

#train_id <- sample(1:nrow(train.data), size = floor(0.8 * nrow(train.data)), replace=FALSE) 

# Split in training and validation (80/20)
validation.data <- data.matrix(xgboost_data %>% filter(yearSeason<=2008 & yearSeason >2002) %>% select(numeric_names))
validation.data <- validation.data[,-1]
validation.label = (current_train %>% filter(yearSeason<=2008 & yearSeason>2002))$isMVP

#train.data <- train.data[train_id,]
#train.label = current_train[train_id,]$isMVP


test.data = data.matrix(xgboost_data %>% filter(yearSeason>2008) %>% select(numeric_names) %>% select_if(is.numeric))
test.data <- test.data[, -1]
test.label = current_test$isMVP
```

## Create xgb.Matrix objects
```{r}
# Transform the two data sets into xgb.Matrix
xgb.train = xgb.DMatrix(data=train.data,label=train.label)
xgb.validation = xgb.DMatrix(data=validation.data,label=validation.label)
xgb.test = xgb.DMatrix(data=test.data,label=test.label)
```

## Define the main parameters
```{r}
# Define the parameters for regression
#eta - 0.2 or 0.3
#gamma ~ 3
#max depth = 6
params = list(
  booster="gbtree",
  eta=0.3,
  objective="binary:logistic",
  eval_metric="logloss",
  eval_metric="error",
  eval_metric="auc",
  scale_pos_weight = 393.6071, 
  gamma = 4,
  max_depth = 6,
  min_child_weight = 10
)
```


## Train the model
```{r}
# Train the XGBoost model
xgb.fit=xgb.train(
  params=params,
  data=xgb.train,
  nrounds=100,
  nthreads=1,
  early_stopping_rounds=10,
  watchlist=list(train=xgb.train,val=xgb.validation),
  verbose=0
)

# Review the final model and results
xgb.fit
importance_matrix <- xgb.importance(colnames(train.data), model = xgb.fit)
importance_matrix_top_features <- importance_matrix[1:20, 1]
xgb.plot.importance(xgb.importance(colnames(train.data), model = xgb.fit))

```

```{r}
# Predict outcomes with the validation data (to tune model)
xgb.pred = predict(xgb.fit, validation.data ,reshape=T)
pred.label = as.factor(round(xgb.pred))
confusionMatrix(as.factor(validation.label), pred.label)

xgb.pred = as.data.frame(xgb.pred)
validation.data.df <- cbind(validation.data, validation.label)
validation.data.df <- cbind(validation.data.df, xgb.pred)
validation.data.df <- cbind(validation.data.df, year2)

#evaluate "accuracy"
validation.data.df <- as.data.frame(validation.data.df)
validation.data.df %>% select(xgb.pred, validation.label, year2) %>% group_by(year2) %>% slice_max(order_by = xgb.pred, n = 1)


# Predict outcomes with the test data (to evaluate results but this MUST BE INDEPENDENT so I can't use these results to tune model)
xgb.pred = predict(xgb.fit,test.data,reshape=T)
pred.label = as.factor(round(xgb.pred))
confusionMatrix(as.factor(test.label), pred.label)

xgb.pred = as.data.frame(xgb.pred)
test.data.df <- cbind(test.data, test.label)
test.data.df <- cbind(test.data.df, xgb.pred)
test.data.df <- cbind(test.data.df, year)

#evaluate "accuracy"
test.data.df <- as.data.frame(test.data.df)
test.data.df %>% select(xgb.pred, test.label, year) %>% group_by(year) %>% slice_max(order_by = xgb.pred, n = 1)
```

# XGBoost (regression)
advanced stats (WS if in, VORP if WS is removed) take over and are the only variable included because of gain of 1

## Training, validation, and test set creation
```{r}
set.seed(1)
xgboost_data <- current_stats
label = xgboost_data$award_share
xgboost_data$award_share = NULL
n = nrow(xgboost_data)
year = (xgboost_data %>% filter(yearSeason>2008))$yearSeason
year2 = (xgboost_data %>% filter(yearSeason>2002 & yearSeason<=2008))$yearSeason

names_backward_select <- c("agePlayer", "pct3PRate", "pctFTRate", "pctDRB", "pctAST", "pctUSG", 
    "ratioOWS", "ratioWS", "ratioBPM", "ratioDBPM","ratioVORP", "pctFT", "minutesPerGame", 
    "fgmPerGame", "fg3mPerGame", "fg2aPerGame", "ftaPerGame", "drbPerGame", "astPerGame", 
    "ptsPerGame", "ptsPerGameTeam", "tovPerGameTeam", "pctFGPerGameTeam")


train.data = data.matrix(xgboost_data %>% filter(yearSeason<=2002) %>% select(names_backward_select) %>% select_if(is.numeric))
train.data <- train.data[,-1]
train.label = (current_train %>% filter(yearSeason<=2002))$award_share
train.label2 = (current_train %>% filter(yearSeason<=2002))$isMVP

#train_id <- sample(1:nrow(train.data), size = floor(0.8 * nrow(train.data)), replace=FALSE) 

# Split in training and validation (80/20)
validation.data <- data.matrix(xgboost_data %>% filter(yearSeason<=2008 & yearSeason >2002) %>% select(names_backward_select))
validation.data <- validation.data[,-1]
validation.label = (current_train %>% filter(yearSeason<=2008 & yearSeason>2002))$award_share
validation.label2 = (current_train %>% filter(yearSeason<=2008 & yearSeason>2002))$isMVP

#train.data <- train.data[train_id,]
#train.label = current_train[train_id,]$isMVP


test.data = data.matrix(xgboost_data %>% filter(yearSeason>2008) %>% select(names_backward_select) %>% select_if(is.numeric))
test.data <- test.data[, -1]
test.label = current_test$award_share
test.label2 = current_test$isMVP
```

## Create xgb.Matrix objects
```{r}
# Transform the two data sets into xgb.Matrix
xgb.train = xgb.DMatrix(data=train.data,label=train.label)
xgb.validation = xgb.DMatrix(data=validation.data,label=validation.label)
xgb.test = xgb.DMatrix(data=test.data,label=test.label)
```

## Define the main parameters
```{r}
# Define the parameters for regression
#eta - 0.2 or 0.3
#gamma ~ 3
#max depth = 6
params = list(
  booster="gbtree",
  eta=0.3,
  objective="reg:squarederror",
  eval_metric="rmse",
  eval_metric="mae",
  scale_pos_weight = 393.6071, 
  gamma = 0,
  max_depth = 6,
  min_child_weight = 10
)
```


## Train the model
```{r}
# Train the XGBoost model
xgb.fit=xgb.train(
  params=params,
  data=xgb.train,
  nrounds=100,
  nthreads=1,
  early_stopping_rounds=10,
  watchlist=list(train=xgb.train,val=xgb.validation),
  verbose=0
)

# Review the final model and results
xgb.fit
importance_matrix <- xgb.importance(colnames(train.data), model = xgb.fit)
importance_matrix_top_features <- importance_matrix[1:20, 1]
xgb.plot.importance(xgb.importance(colnames(train.data), model = xgb.fit))

```

```{r}
# Predict outcomes with the validation data (to tune model)
xgb.pred = predict(xgb.fit, validation.data ,reshape=T)
xgb.pred = as.data.frame(xgb.pred)
validation.data.df <- cbind(validation.data, validation.label)
validation.data.df <- cbind(validation.data.df, xgb.pred)
validation.data.df <- cbind(validation.data.df, year2)
validation.data.df <- cbind(validation.data.df, validation.label2)

#evaluate "accuracy"
validation.data.df <- as.data.frame(validation.data.df)
validation.data.df %>% select(year2, xgb.pred, validation.label, validation.label2) %>% group_by(year2) %>% slice_max(order_by = xgb.pred, n = 1)


# Predict outcomes with the test data (to evaluate results but this MUST BE INDEPENDENT so I can't use these results to tune model)
xgb.pred = predict(xgb.fit,test.data,reshape=T)
xgb.pred = as.data.frame(xgb.pred)
test.data.df <- cbind(test.data, test.label)
test.data.df <- cbind(test.data.df, xgb.pred)
test.data.df <- cbind(test.data.df, year)
test.data.df <- cbind(test.data.df, test.label2)

#evaluate "accuracy"
test.data.df <- as.data.frame(test.data.df)
test.data.df %>% select(xgb.pred, test.label, test.label2, year) %>% group_by(year) %>% slice_max(order_by = xgb.pred, n=1)
```

# XGboost (simpler, regression on award share)
```{r}
xgb_train_numeric <- current_train_numeric %>% select(-isMVP, -yearSeason, -year, -merge_year, -pctORB, -ratioDBPM, -pctFT, -pctFG, -fg3aPerGame, -fg3mPerGame, -orbPerGame, -ratioDWS)
nrow(filter(xgb_train_numeric, award_share==0)) / nrow(filter(xgb_train_numeric, award_share != 0))

tune_grid <- expand.grid(
  nrounds = seq(from = 200, to = 500, by = 100),
  eta = c(0.025, 0.05, 0.1, 0.3),
  max_depth = 2:10,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

tune_control <- caret::trainControl(
  method = "cv", # cross-validation
  number = 10, # with n folds 
  #index = createFolds(tr_treated$Id_clean), # fix the folds
  verboseIter = TRUE, # no training log
  allowParallel = FALSE # FALSE for reproducible results 
)

award_share_tuned_mod <- xgb_train_numeric %>%
  caret::train(
  award_share ~ .,
  data = .,
  trControl = tune_control,
  tuneGrid = tune_grid,
  method = "xgbTree",
  verbose = FALSE,
  scale_pos_weight = 16.5497
)

#importance tables and graphs
imp <- varImp(award_share_tuned_mod, scale = F)
sum(imp$importance$Overall)
#imp <- imp %>% mutate(Overall = format(round(Overall, 5), scientific = FALSE, big.mark = ","))

ggplot(imp, mapping=NULL, top=10) +
  labs(title = "XGBoost Feature Importance") +
  theme(plot.title = element_text(hjust = 0.5, color = "Purple"), plot.subtitle = element_text(hjust = 0.5, color = "Salmon"))

imp2 <- imp$importance %>% rownames_to_column()
ggplot(imp2[1:10,], mapping=aes(x=Overall, y=reorder(rowname, Overall))) +
  geom_col() +
  labs(title = "XGBoost Feature Importance", x="Importance", y="Variable") +
  theme(plot.title = element_text(hjust = 0.5, color = "Purple"), plot.subtitle = element_text(hjust = 0.5, color = "Salmon")) +
  theme_clean() + 
  scale_y_discrete(labels = c("AST%", "FG2A/G", "FG2M/G", "AST/G", "TOV/G", "FGA/G", "PER", "VORP", "Win%", "WS"))

current_test_complete <- current_test %>% 
  select(-isMVP, -yearSeason, -year, -merge_year) %>%
  na.omit()
award_share_preds <- predict(award_share_tuned_mod, newdata = current_test_complete, type = "raw")
current_test_results <- cbind(current_test_complete, award_share_preds)
current_test_results<- current_test_results %>% rename("pred" = "...66")

current_test_results %>%
  select(namePlayer, slugSeason.x, award_share, pred) %>%
  arrange(slugSeason.x, -pred) %>%
  group_by(slugSeason.x) %>%
  filter(row_number() %in% c(1))

current_test_results %>%
  select(namePlayer, slugSeason.x, award_share, pred) %>%
  arrange(slugSeason.x, -award_share) %>%
  group_by(slugSeason.x) %>%
  filter(row_number() %in% c(1))


award_share_tuned_mod$results
award_share_tuned_mod$bestTune

sqrt(mean((current_test_results$award_share - current_test_results$pred)^2))
sqrt(mean((current_pred$award_share - current_pred$pred)^2, na.rm = T))
#saveRDS(xgb_attack_angle_tuned_mod, "attack_angle_mod.rds")

#xgb_attack_angle_tuned_mod <- readRDS("attack_angle_mod.rds")

```

# XGBoost (simpler, classification)
```{r}
xgb_train_numeric <- current_train_numeric %>% select(-award_share, -yearSeason, -year, -merge_year)
xgb_train_numeric$isMVP <- as.factor(xgb_train_numeric$isMVP)


tune_grid <- expand.grid(
  nrounds = seq(from = 200, to = 500, by = 100),
  eta = c(0.025, 0.05, 0.1, 0.3),
  max_depth = 2:10,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

tune_control <- caret::trainControl(
  method = "cv", # cross-validation
  number = 3, # with n folds 
  #index = createFolds(tr_treated$Id_clean), # fix the folds
  verboseIter = TRUE, # no training log
  allowParallel = FALSE # FALSE for reproducible results 
)

isMVP_tuned_mod <- xgb_train_numeric %>%
  caret::train(
  isMVP ~ .,
  data = .,
  trControl = tune_control,
  tuneGrid = tune_grid,
  method = "xgbTree",
  verbose = FALSE
)

#importance tables and graphs
imp <- varImp(award_share_tuned_mod, scale = F)
sum(imp$importance$Overall)
#imp <- imp %>% mutate(Overall = format(round(Overall, 5), scientific = FALSE, big.mark = ","))

ggplot(imp, mapping=NULL, top=10) +
  labs(title = "XGBoost Feature Importance") +
  theme(plot.title = element_text(hjust = 0.5, color = "Purple"), plot.subtitle = element_text(hjust = 0.5, color = "Salmon")) 

current_test_complete <- current_test %>% 
  select(-award_share, -yearSeason, -year, -merge_year) %>%
  na.omit()
award_share_preds <- predict(isMVP_tuned_mod, newdata = current_test_complete, type = "prob")
current_test_results <- cbind(current_test_complete, award_share_preds)

current_test_results %>%
  select(namePlayer, slugSeason.x, isMVP, `0`, `1`) %>%
  arrange(slugSeason.x, -`1`) %>%
  group_by(slugSeason.x) %>%
  filter(row_number() %in% c(1))

current_test_results %>%
  select(namePlayer, slugSeason.x, isMVP, `0`, `1`) %>%
  arrange(slugSeason.x, -isMVP) %>%
  group_by(slugSeason.x) %>%
  filter(row_number() %in% c(1))
```


# Exporting Model Data
```{r}
write_csv(lagged_stats_model_likely, "model_data.csv")
write_csv(sim_train, "sim_train.csv")
write_csv(current_train, "current_train2.csv")
write_csv(current_test, "current_test2.csv")

htmlTable(evaluation)

```

# New Season Data to Predict On
```{r}
#advanced and per game stats
z_player_stats <- bref_players_stats(seasons = 2023, tables = c("advanced", "per_game"))
z_player_stats <- z_player_stats %>% group_by(namePlayer, slugSeason) %>% dplyr::slice(1)

#what draft pick they were
#need more identifiers to prevent duplicates
z_drafts <- drafts(draft_years = 2022, nest_data = FALSE, return_message = TRUE)
z_drafts <- z_drafts %>% select(namePlayer, numberPickOverall, idPlayer)
z_drafts <- rbind(drafts2, z_drafts)

z_drafts %>% group_by(namePlayer) %>% summarise(n= n()) %>% filter(n>1)

z_player_stats_draft_pick <- left_join(z_player_stats, z_drafts, by=c("idPlayerNBA"="idPlayer", "namePlayer"))


#adding attributes to main data
z_full_player_stats <- left_join(z_player_stats_draft_pick, attributes, by=c("namePlayer"="Player"))

#prior MVPs variables
total_mvps_2023 <- data.frame(namePlayer = c("LeBron James", "Giannis Antetokounmpo", "Nikola Jokic", "James Harden", "Russell Westbrook", "Stephen Curry", "Kevin Durant", "Derrick Rose"), totalMVP = c(4, 2, 2, 1, 1, 2, 1, 1))
z_full_player_stats <- z_full_player_stats %>% left_join(total_mvps_2023)
z_full_player_stats <- z_full_player_stats %>% 
  mutate(totalMVP = ifelse(is.na(totalMVP), 0, totalMVP)) 
  
#removing unnecessary variables 
z_full_player_stats_clean <- z_full_player_stats %>% select(c("slugSeason", "namePlayer", "groupPosition", "yearSeason", 
"slugPosition", "agePlayer", "slugTeamBREF", "countGames", "slugTeamsBREF", "ratioPER", 
"pctTrueShooting", "pct3PRate", "pctFTRate", "pctORB", "pctDRB", 
"pctTRB", "pctAST", "pctSTL", "pctBLK", "pctTOV", "pctUSG", "ratioOWS", 
"ratioDWS", "ratioWS", "ratioWSPer48", "ratioOBPM", "ratioDBPM", 
"ratioBPM", "ratioVORP", "countGamesStarted", 
"pctFG", "pctFG3", "pctFG2", "pctEFG", "pctFT", "minutesPerGame", 
"fgmPerGame", "fgaPerGame", "fg3mPerGame", "fg3aPerGame", "fg2mPerGame", 
"fg2aPerGame", "ftmPerGame", "ftaPerGame", "orbPerGame", "drbPerGame", 
"trbPerGame", "astPerGame", "stlPerGame", "blkPerGame", "tovPerGame", 
"pfPerGame", "ptsPerGame", "numberPickOverall", "height", "weight", "totalMVP"))

z_team_stats <- data.frame()
for (i in 2023) {
  df <- bref_teams_stats(seasons = i, assign_to_environment = F)
  df <- df[[2]][[1]]
  df <- df %>% select(slugSeason, yearSeason, slugTeamBREF, pctWins, ptsPerGameTeam, astPerGameTeam, drbPerGameTeam, tovPerGameTeam, pctFGPerGameTeam, pctFTPerGameTeam)
  z_team_stats <- df
}

#different from training data in that its missing isMVP, award share, year, merge year (hopefully won't matter)
z_team_player_stats <- left_join(z_full_player_stats_clean, z_team_stats, by=c("slugTeamBREF", "yearSeason"))
z_team_player_stats <- z_team_player_stats %>% mutate(numberPickOverall = ifelse(is.na(numberPickOverall), 0, numberPickOverall))

#write.csv(z_team_player_stats, "MVP_predictions_2023.csv")

```

# 2022-23 Predictions
```{r}
attributes_current <- readxl::read_excel("height_weight_current.xlsx")
attributes_current <- attributes_current %>%
  mutate(height = (as.numeric(substr(ht, 1, 1))*12 + as.numeric(word(ht, 2, sep="-"))) * 2.54,
         weight = wt*0.454) %>%
  select(-ht, -wt)

z_team_player_stats_full <- left_join(z_team_player_stats, attributes_current, by=c("namePlayer"="Player"))
z_team_player_stats_full <- z_team_player_stats_full %>%
  select(-height.x, -weight.x) %>%
  rename("height" = "height.y", "weight" = "weight.y") %>%
  filter(!is.na(height))

na_count <- sapply(z_team_player_stats_full, function(y) sum(length(which(is.na(y)))))
na_count <- data.frame(na_count)

z_team_player_stats_full <- z_team_player_stats_full %>% filter(!is.na(pctWins))

new_preds <- predict(award_share_tuned_mod, newdata = z_team_player_stats_full, type = "raw")
new_preds <- cbind(z_team_player_stats_full, new_preds)
new_preds <- new_preds %>% rename("pred" = "...66")

new_preds %>%
  select(namePlayer, pred, ratioWS, ratioVORP, pctWins, ratioPER) %>%
  arrange(-pred)

htmlTable(new_preds %>% filter(minutesPerGame > 20) %>% select(namePlayer, pred)%>% arrange(-pred))

```

